---
title: "Exercise 3 Amr Maraqa"
output: 
  md_document: default
  pdf_document: default
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all") })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(here)
library(arrow)
library(gender)
library(lubridate)

library(tidyverse)
library(igraph)
library(tidygraph)
library(ggraph)
library(gridExtra)
```

## Load data

Load the following data:
  + applications from `app_data_sample.parquet`
  + edges from `edges_sample.csv`

```{r load-data, echo=FALSE}
# change to your own path!
data_path <- "C:\\Users\\m.maraqa\\Desktop\\MMA courses\\Winter 2023\\Organizational Network Analysis\\Exercise 3\\"
applications <- read_parquet(paste0(data_path,"app_data_sample.parquet"))
edges <- read_csv(paste0(data_path,"edges_sample.csv"))

applications
edges
```

## Get gender for examiners

We'll get gender based on the first name of the examiner, which is recorded in the field `examiner_name_first`. We'll use library `gender` for that, relying on a modified version of their own [example](https://cran.r-project.org/web/packages/gender/vignettes/predicting-gender.html).

Note that there are over 2 million records in the applications table -- that's because there are many records for each examiner, as many as the number of applications that examiner worked on during this time frame. Our first step therefore is to get all *unique* names in a separate list `examiner_names`. We will then guess gender for each one and will join this table back to the original dataset. So, let's get names without repetition:

```{r gender-1, echo=FALSE}
library(gender)
#install_genderdata_package() # only run this line the first time you use the package, to get data for it

# get a list of first names without repetitions
examiner_names <- applications %>% 
  distinct(examiner_name_first)

examiner_names
```

Now let's use function `gender()` as shown in the example for the package to attach a gender and probability to each name and put the results into the table `examiner_names_gender`

```{r gender-2, echo=FALSE}
# get a table of names and gender
examiner_names_gender <- examiner_names %>% 
  do(results = gender(.$examiner_name_first, method = "ssa")) %>% 
  unnest(cols = c(results), keep_empty = TRUE) %>% 
  select(
    examiner_name_first = name,
    gender,
    proportion_female
  )

examiner_names_gender
```

Finally, let's join that table back to our original applications data and discard the temporary tables we have just created to reduce clutter in our environment.

```{r gender-3, echo=FALSE}
# remove extra colums from the gender table
examiner_names_gender <- examiner_names_gender %>% 
  select(examiner_name_first, gender)

# joining gender back to the dataset
applications <- applications %>% 
  left_join(examiner_names_gender, by = "examiner_name_first")

# cleaning up
rm(examiner_names)
rm(examiner_names_gender)
gc()

```


## Guess the examiner's race

We'll now use package `wru` to estimate likely race of an examiner. Just like with gender, we'll get a list of unique names first, only now we are using surnames.

```{r race-1, echo=FALSE}
library(wru)

examiner_surnames <- applications %>% 
  select(surname = examiner_name_last) %>% 
  distinct()

examiner_surnames
```
We'll follow the instructions for the package outlined here [https://github.com/kosukeimai/wru](https://github.com/kosukeimai/wru).

```{r race-2, echo=FALSE}
examiner_race <- predict_race(voter.file = examiner_surnames, surname.only = T) %>% 
  as_tibble()

examiner_race
```

As you can see, we get probabilities across five broad US Census categories: white, black, Hispanic, Asian and other. (Some of you may correctly point out that Hispanic is not a race category in the US Census, but these are the limitations of this package.)

Our final step here is to pick the race category that has the highest probability for each last name and then join the table back to the main applications table. See this example for comparing values across columns: [https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/](https://www.tidyverse.org/blog/2020/04/dplyr-1-0-0-rowwise/). And this one for `case_when()` function: [https://dplyr.tidyverse.org/reference/case_when.html](https://dplyr.tidyverse.org/reference/case_when.html).

```{r race-3, echo=FALSE}
examiner_race <- examiner_race %>% 
  mutate(max_race_p = pmax(pred.asi, pred.bla, pred.his, pred.oth, pred.whi)) %>% 
  mutate(race = case_when(
    max_race_p == pred.asi ~ "Asian",
    max_race_p == pred.bla ~ "black",
    max_race_p == pred.his ~ "Hispanic",
    max_race_p == pred.oth ~ "other",
    max_race_p == pred.whi ~ "white",
    TRUE ~ NA_character_
  ))

examiner_race
```

Let's join the data back to the applications table.

```{r race-4, echo=FALSE}
# removing extra columns
examiner_race <- examiner_race %>% 
  select(surname,race)

applications <- applications %>% 
  left_join(examiner_race, by = c("examiner_name_last" = "surname"))

rm(examiner_race)
rm(examiner_surnames)
gc()
```


## Examiner's tenure 

To figure out the timespan for which we observe each examiner in the applications data, let's find the first and the last observed date for each examiner. We'll first get examiner IDs and application dates in a separate table, for ease of manipulation. We'll keep examiner ID (the field `examiner_id`), and earliest and latest dates for each application (`filing_date` and `appl_status_date` respectively). We'll use functions in package `lubridate` to work with date and time values.

```{r tenure-1, echo=FALSE}
library(lubridate) # to work with dates

examiner_dates <- applications %>% 
  select(examiner_id, filing_date, appl_status_date) 

examiner_dates
```

The dates look inconsistent in terms of formatting. Let's make them consistent. We'll create new variables `start_date` and `end_date`.

```{r tenure-2, echo=FALSE}
examiner_dates <- examiner_dates %>% 
  mutate(start_date = ymd(filing_date), end_date = as_date(dmy_hms(appl_status_date)))
```

Let's now identify the earliest and the latest date for each examiner and calculate the difference in days, which is their tenure in the organization.

```{r tenure-3, echo=FALSE}
examiner_dates <- examiner_dates %>% 
  group_by(examiner_id) %>% 
  summarise(
    earliest_date = min(start_date, na.rm = TRUE), 
    latest_date = max(end_date, na.rm = TRUE),
    tenure_days = interval(earliest_date, latest_date) %/% days(1)
    ) %>% 
  filter(year(latest_date)<2018)%>%
  mutate(tenure_years = tenure_days / 365) %>%
    mutate(tenure = case_when(
      tenure_years <= 1 ~ '<1',
      tenure_years <= 2 ~ '1-2',
      tenure_years <= 5 ~ '3-5',
      tenure_years <= 9 ~ '6-9',
      tenure_years <= 14 ~ '10-14',
      tenure_years <= 100 ~ '15+',
      TRUE ~ NA_character_
    ))

examiner_dates
```

Joining back to the applications data.

```{r tenure-4, echo=FALSE}
applications <- applications %>% 
  left_join(examiner_dates, by = "examiner_id")

rm(examiner_dates)
gc()
```

Creating work group column and dropping NAs
```{r preprocessing, echo=FALSE}
#Creating work group column
applications$examiner_workgroup <- as.integer(applications$examiner_art_unit/10)
applications$examiner_workgroup <- as.character(applications$examiner_workgroup)

#dropping NA values in gender
applications <- applications %>%
  drop_na(gender, race, tenure_days)
```

Finding the work group numbers of each art unit and creating a data set for work groups 213 and 174
```{r 2 work groups, echo=FALSE}
# Examiner Data
examiner_data <- applications %>%
  distinct(examiner_id, examiner_gender = gender, 
           examiner_race = race, examiner_tenure = tenure) 
# Work Groups
examiner_subset <- applications %>% 
  filter(examiner_workgroup %in% c('213', '174')) %>%
  distinct(examiner_id, examiner_workgroup) %>%
  left_join(examiner_data, by='examiner_id')

num_examiners <- examiner_subset %>% 
  count(examiner_workgroup)
```

Getting summary statistics for work groups
```{r wg summary statistics, echo=FALSE}
t_gend <- examiner_subset %>% count(examiner_workgroup, examiner_gender) %>%
  group_by(examiner_workgroup) %>% mutate(freq = n / sum(n) * 100) %>%
  select(examiner_workgroup, examiner_gender, freq) %>% 
  mutate(freq = round(freq, 2)) %>%
  pivot_wider(names_from = examiner_gender, values_from = freq)
t_race <- examiner_subset %>% count(examiner_workgroup, examiner_race) %>%
  group_by(examiner_workgroup) %>% mutate(freq = n / sum(n) * 100) %>%
  select(examiner_workgroup, examiner_race, freq) %>% 
  mutate(freq = round(freq, 2)) %>%
  pivot_wider(names_from = examiner_race, values_from = freq)
t_tenure <- examiner_subset %>% count(examiner_workgroup, examiner_tenure) %>%
  group_by(examiner_workgroup) %>% mutate(freq = n / sum(n) * 100) %>%
  mutate(freq = round(freq, 2)) %>%
  select(examiner_workgroup, examiner_tenure, freq) %>% 
  pivot_wider(names_from = examiner_tenure, values_from = freq)

```

```{r summary plots, echo=FALSE}
p_gend <- ggplot(examiner_subset, aes(x=examiner_gender, y=..prop.., 
                                      fill=examiner_workgroup, 
                                      group=examiner_workgroup)) +
  geom_bar(aes(), stat='count', position='dodge') +
  scale_y_continuous(labels = scales::percent_format())
p_race <- ggplot(examiner_subset, aes(x=examiner_race, y=..prop.., 
                                      fill=examiner_workgroup, 
                                      group=examiner_workgroup)) +
  geom_bar(aes(), stat='count', position='dodge') +
  scale_y_continuous(labels = scales::percent_format())
p_tenure <- ggplot(examiner_subset, aes(x=examiner_tenure, y=..prop.., 
                                        fill=examiner_workgroup, 
                                        group=examiner_workgroup)) +
  geom_bar(aes(), stat='count', position='dodge')

```

```{r Create Network, echo=FALSE}
edge_subset <- edges %>% 
  filter(ego_examiner_id %in% examiner_subset$examiner_id &
           alter_examiner_id %in% examiner_subset$examiner_id) %>%
  drop_na() %>% 
  select(to = ego_examiner_id, from = alter_examiner_id)
node_subset <- edge_subset %>% 
  pivot_longer(cols=c('from','to')) %>% 
  distinct(examiner_id = value) %>%
  left_join(examiner_data, on='examiner_id') %>%
  distinct(examiner_id, examiner_gender, examiner_race, examiner_tenure) %>%
  rename(name = examiner_id) %>%
  mutate(name = as.character(name))
network <- graph_from_data_frame(edge_subset, directed = TRUE) %>%
  as_tbl_graph() %>%
  left_join(node_subset, by='name')
```

```{r Network Metrics, echo=FALSE}
network <- network %>%
  mutate(degree = centrality_degree(),
         betweenness = centrality_betweenness()) %>%
  mutate(avg = (degree + betweenness)/2) %>%
  mutate(label = paste0(name, '\n',
                        'Degree: ',round(degree,2), '\n',
                        'Betweenness: ',round(betweenness,2), '\n',
                        'Avg: ',round(avg,2)))
```

```{r Plot Network, echo=FALSE}
set.seed(1)
net_gender <- network %>%
  ggraph(layout="mds") +
  geom_edge_link(edge_colour = "#d3d3d3", alpha=0.1) +
  geom_node_point(aes(color=examiner_gender, size=avg)) +
  theme_void()
set.seed(1)
net_race <- network %>%
  ggraph(layout="mds") +
  geom_edge_link(edge_colour = "#d3d3d3", alpha=0.1) +
  geom_node_point(aes(color=examiner_race, size=avg)) +
  theme_void()

```
```{r Discussion, echo=FALSE}

network_data <- network %>% as.data.frame() %>% as_tibble()
disc_gend_mean <- network_data %>% 
  group_by(examiner_gender) %>%
  summarize(mean_degree = mean(degree), 
            mean_bet = mean(betweenness))
disc_gend_top_degree <- network_data %>% 
  arrange(desc(degree)) %>%
  group_by(examiner_gender) %>%
  top_frac(0.1, degree) %>%
  summarize(top10_degree = mean(degree)) %>%
  mutate(top10_degree = round(top10_degree, 2))
disc_gend_top_bet <- network_data %>% 
  arrange(desc(betweenness)) %>%
  group_by(examiner_gender) %>%
  top_frac(0.1, betweenness) %>%
  summarize(top10_bet = mean(betweenness)) %>%
  mutate(top10_bet = round(top10_bet, 2))
disc_gend_top <- disc_gend_top_degree %>% 
  left_join(disc_gend_top_bet, on='examiner_gender')
disc_gend <- disc_gend_top %>% 
  left_join(disc_gend_mean, on='examiner_gender')
disc_race_mean <- network_data %>% 
  group_by(examiner_race) %>%
  summarize(mean_degree = mean(degree), 
            mean_bet = mean(betweenness))
disc_race_top_degree <- network_data %>% 
  arrange(desc(degree)) %>%
  group_by(examiner_race) %>%
  top_frac(0.1, degree) %>%
  summarize(top10_degree = mean(degree)) %>%
  mutate(top10_degree = round(top10_degree, 2))
disc_race_top_bet <- network_data %>% 
  arrange(desc(betweenness)) %>%
  group_by(examiner_race) %>%
  top_frac(0.1, betweenness) %>%
  summarize(top10_bet = mean(betweenness)) %>%
  mutate(top10_bet = round(top10_bet, 2))
disc_race_top <- disc_race_top_degree %>% 
  left_join(disc_race_top_bet, on='examiner_race')
disc_race <- disc_race_top %>% 
  left_join(disc_race_mean, on='examiner_race')
```

## Summary Statistics

Work groups 213 and 174 are evaluated. 

- Work group 213 has 262 employees, and 174 has 252
- Work group 213 is more male dominated with 78.2% vs 65.5% for 174.
- Both work groups consist predominantly of white examiners, with Asian examiners constituting the second biggest race group.
- The average tenure among examiners in Work group 213 is lower than that of the examiners in 174. This means that 213 has bigger groups of less experienced examiners, which may also mean that examiners in 213 are younger, on average.

```{r echo=FALSE}
kable(num_examiners, caption='Total Number of Examiners')
kable(t_gend, caption='Gender Distribution')
kable(t_race, caption='Race Distribution')
kable(t_tenure, caption='Tenure Distribution')
grid.arrange(p_gend, p_race, p_tenure)
```


## Network Visualization

There seems to be three groups which are connected through one node. This is critical, as the absence of this examiner, would render communication between these clusters impossible. Given that these clusters originate from two work groups, it can deduced that there is a clear separation between examiners in one of them, which can render it dysfunctional should the connecting examiner leave. This is a sign to the the USPTO that teamwork within and between these two work groups is at jeopardy due to the centralization of the communication. 

Within each of the 3 clusters, no segregation by gender or by race appears to exist. This could be due to the non-dominant groups being too small to form their own cluster, to the fact that the employees are interested in maintaining diverse groups.

```{r echo=FALSE}
net_gender
net_race
```

## Discussion
Gender:
The difference in gender doesn't seem to impact the popularity of an examiner in and across the 213 and 174 work groups; the measures of degree centrality and betweenness are very similar for both genders.

Race:
On the other hand, there exists a clear discrepancy between the various races within the work groups. White people hold the most pivotal positions within the network and are most frequently the channels of communication between examiners. Meanwhile, black people seem to always be on the ends of the networks or aren't detrimental to their flow of information. As for popularity, White and Asian examiners are very similar. It's hard to attribute these differences to racial discrimination, however, as the white community is over-represented in these work groups. The problem may be in the systematic racism in the recruitment process of the UPSTO, but no conclusions can be drawn about the examiners and the relationships between them.  

```{r echo=FALSE}
kable(disc_gend, caption='Gender Centrality Scores')
kable(disc_race,xcaption='Race Centrality Scores')
```
